Recovery of mountain plant communities in response to reductions in Nitrogen emissions is hidden by other drivers of global change

### 2018-06-14 Manuscript.Rmd
Teil der Diskussion geschrieben und Referenzen ergänzt.

@@

@@

This temporal stability as inferred from the community measures was, however, in contrast to a rather large observed temporal turnover of species. The average percentage $\pm$ SD of species that differed between the first and second survey at a site was `r mean(100 * sites$change1 / (sites$nochange1 + sites$change1))` $\pm$ `r sd(100 * sites$change1 / (sites$nochange1 + sites$change1))`% and the percentage of species that differ between the second and third survey was `r mean(100 * sites$change2 / (sites$nochange2 + sites$change2))` $\pm$ `r sd(100 * sites$change2 / (sites$nochange2 + sites$change2))`%. Thus, it seemed that the turnover from the first/second survey to the turnover of the second/third survey moderately decreased (90% Credible interval of the change in turnover estimated from the Binomial generalized linear mixed model: `r CrI`). Variation in species turnover was largest along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition (Table \@ref(tab:turnovertab)). The other three gradients were less important to explain the variation in species turnover among sites.

@@

# Discussion {-}
- *General points*: Although N deposition considerably declined between 2005 and 2015, we could not detect major shifts in plant community structure during the same time period. 
Although NOx emissions in Switzerland decreased by 46% between 1990 and 2010 and NH3 emissions by 14% (Maas and Grennfelt, 2016), nitrogen deposition on our plots decreased by only 2.7 kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015 (Meteotest, 2018). This is only about one tenth of the decrease in England in the same period, where Nitrogen deposition decreased by 24kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 (Storkey et al. 2015). In 2015, average N deposition was 14.8 kg ha$^{-1}$ yr$^{-1}$, about 85% of 2000 deposition, and due to the rather short period of time and the small decrease in nitrogen deposition, it is not surprising that species richness has not reacted significantly.

- *Replacement of oligotrophic with eutrophic species is faster than the oposite direction*: Eutrophic species have rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high N deposition. This suggests that it takes 
more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. Climatic effects may be more likely to be reversed than effects due to fertilization.

- *Methodological point*: The rather large spatial turnover might be partly explained by species that remained undetected in one of the surveys. From the randomly selected quality control of the BDM 17 mountain hay meadows were investigated from 2003 to 2017. Species richness was surveyed in the same year by two independent botanists. The species match is (mean $\pm$ SD) 85 $\pm$ 5.2 per cent, suggesting that at least a portion of the turn over is due to agent effects. However, our results suggest that turnover is caused at least partly by species with specific indicator values. This deviation from what we would expect under random species turnover is unlikely to be explained by species that remained undetected.

@@

- *Space for time substitution*: Often observational studies infer the change of plant diversity along a gradient of N deposi- tion. Thus, they infer how the spatial variation in species richness is related to N deposition and assume that this spatial variation in species richness arose because over time some areas lost more species than others because they chronically experienced higher N deposition. Alt- hough there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of N deposition on plant diversity (Stevens et al. 2010), they can not replace stud- ies that relate temporal patterns in species with N deposition (De Schrijver et al. 2011). While recovery of acidified surface waters has been well investigated (De Vries et al. 2015), there are only a limited number of studies inferring temporal trends of plant species diversity relat- ed to varying amounts of N-deposition. Storkey et al. (2015) demonstrated a positive response of biodiversity to reducing N addition from either atmospheric pollution or fertilizers in the Park Grass Experiment: «The proportion of legumes, species richness and diversity increased across the experiment between 1991 and 2012 as N-deposition declined». For forest floor vegetation in permanent plots across Europe the exceedance of critical loads of N over a peri- od from 9 to 42 years had negative effects on the cover of oligotrophic plant species, i.e spe- cies that prefer nutrient-poor soils, although species richness remained constant (Dirnböck et al. 2014). Another example of recovery in eutrophicated habitats gives the recovery of species richness in previously fertilized plots (Clark and Tilman 2008). In this study, the recorded recovery in species richness within one or two decade was likely due to the species rich vege- tation surrounding the experimental plots, from where immigration was easily feasible.

# Conclusions {-}
xxx

>>>>

NOx emissions in Switzerland decreased by 46% between 1990 and 2010 and NH3 emissions by 14% (Maas and Grennfelt, 2016)

@@

To test data quality, independent replicate surveys are routinely performed within the BDM programm by botansists who were not involved in the regular BDM surveys. The regular surveyors did not know if and which sites were replicated [@Plattner2004]. We used the data from `r nrow(turnover_double)` such replicated surveys that were available to calculate the pseudo-turnover, which is the proportion of species that differed between two surveys that were conducted by two different surveyors during the same year on the same site.

@@

The average $\pm$ SD percentage of species that differed between replicated surveys was `r mean(100 * turnover_double$turnover)` $\pm$ `r sd(100 * turnover_double$turnover)`. This pseud-turnover was lower than the observed temporal turnover: the average percentage of species that differed between the first and second survey at a site was `r mean(100 * sites$change1 / (sites$nochange1 + sites$change1))` $\pm$ `r sd(100 * sites$change1 / (sites$nochange1 + sites$change1))`% and the percentage of species that differ between the second and third survey was `r mean(100 * sites$change2 / (sites$nochange2 + sites$change2))` $\pm$ `r sd(100 * sites$change2 / (sites$nochange2 + sites$change2))`%. Thus, it seemed that the turnover from the first/second survey to the turnover of the second/third survey moderately decreased (90% Credible interval of the change in turnover estimated from the Binomial generalized linear mixed model: `r CrI`). Variation in species turnover was largest along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition (Table \@ref(tab:turnovertab)). The other three gradients were less important to explain the variation in species turnover among sites.

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could not detect major shifts in plant community structure during the same time period (Table \@ref(tab:communitytrendstab)). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperature, the constant aveage nutrients value suggests that the decrease in Nitrogen deposition did not yet affected plant communities. However, the stability in community structure was accompanied by marked turnover in species identities. Species turnover especially varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition. These results are in line with recent studies suggesting that species loss (or gain) is not necessarily apparent in assessments of contemporary communities [@Vellend2013, @Dornelas2014, @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012, @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive change in species communitis could be inferred from comparing the charactereistics of disappearing or colonizign species with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site. The nutrient values of colonization species showed larges deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of the soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England in the same period, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This might explain why we observed no change in average nutrient value of communities. An other explanation might be that communtiy change due to the change in the rate of Nitrogen deposition was overruled by other anthropogenic pressures including climate change. In this study eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at high N deposition. This suggests that it takes more time to replace eutrophic by oligotrophic species than replacing oligotrophic by eutrophic species. This is in contrast to the species' indicator values for temperature: species with low and high temperature values had both optimal temperatures were colonization and local survival probabilities were high (see Fig. . Thus comunitiy change in response to climate warming might be fast because low temperature species have lower local survival and high temperature species have higher colonization probability.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition. They assume that the spatial variation in species richness arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of N-deposition [@Storkey2015]. In @Clark2008, the recorded recovery in species richness within one or two decade was likely due to the species rich vegetation surrounding the experimental plots, from where immigration was easily feasible. 
This is in line with our results sugggesting that it is mainly colonizing and not disappearing species are important to shift back from euttrophic to oligotrophic communities.

- *Methological issue*: The rather large spatial turnover could be partly explained by species that remained undetected in one of the surveys. However, our results suggest that turnover is caused at least partly by species with specific indicator values. This deviation from what we would expect under random species turnover is unlikely to be explained by species that remained undetected.

@@

# Conclusions {-}
Our study adds to the understanding of contemporary biodiversity change that is currently not not well understood (Magurran et al 2018.pdf). Our results support the notion of Hillebrand_et_al-2018.pdf that biodiversity monitoring programmes that quantify species turnover will generate a far more reliable understanding of the biotic response to changing environments than programmes solely tracking species richness.
